Symbiont shuffling linked to differential photochemical dynamics of *Symbiodinium*

### 2017-07-18 manuscript/manuscript.Rmd
RS edits

---
title: Symbiont shuffling linked to differential photochemical dynamics of *Symbiodinium*
  in three Caribbean reef corals
author:
- name: Ross Cunning
  affiliation: MBE
  footnote: Corresponding Author
  email: ross.cunning@gmail.com
- name: Rachel N. Silverstein
  affiliation: MBE,MWK
- name: Andrew C. Baker
  affiliation: MBE

bibliography: library.bib
output:
  rticles::elsevier_article:
    fig_caption: yes
  word_document: null
csl: coral-reefs.csl
address:
- address: Department of Marine Biology and Ecology, Rosenstiel School of Marine and
    Atmospheric Science, University of Miami, 4600 Rickenbacker Causeway, Miami, FL
    33149, USA
  code: MBE
- address: Miami Waterkeeper, 12568 N Kendall Drive, Miami, FL 33185, USA
  code: MWK
---
```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(lmerTest)
library(MASS)
```

@@

# Abstract
  Dynamic symbioses with functionally diverse dinoflagellate algae in the genus *Symbiodinium* may allow some reef corals to alter their phenotypes through 'symbiont shuffling', or changes in symbiont community composition. In particular, corals have been shown to become more bleaching-resistant by increasing the relative abundance of thermally tolerant *Symbiodinium* in clade D after bleaching. Despite the immediate relevance of this phenomenon to corals living in warming oceans -- and to interventions aimed at boosting coral resilience -- the mechanisms governing how, why, and when symbiont shuffling occurs are still poorly understood. Here, we performed controlled thermal bleaching and recovery experiments on three species of Caribbean reef-building corals, and show that the degree of symbiont shuffling is determined by (1) the difference in photochemical efficiency (*F~v~/F~m~*) of co-occurring symbionts under stress, and (2) the duration of stress exposure. The photochemical advantage of D1a (*S. trenchii*) under stress was greatest in *Montastraea cavernosa*, intermediate in *Siderastrea siderea*, and lowest in *Orbicella faveolata*, and correlated positively with the magnitude of increase in D1a. In holobionts where D1a had less of an advantage (i.e., a smaller difference in *F~v~/F~m~*), longer periods of heat stress were required to elicit commensurate increases in D1a. These findings help elucidate general rules governing the process of symbiont shuffling in reef corals, which will help refine predictions of coral responses to environmental change, and inform interventions to manipulate symbiont communities to enhance coral resilience.

@@

Coral reef ecosystems are declining globally at an alarming rate, owing primarily to mass coral bleaching caused by anthropogenic increases in sea surface temperature [@Hughes:2017bs]. Heat-induced bleaching, or the loss of corals' nutritional symbionts (*Symbiodinium* spp.; @Jokiel:1977p7353), often leads to coral mortality, which precipitates reef ecosystem degradation [@Baker:2008p3421].

@@

In both experiments, maximum photochemical efficiency (*F~v~/F~m~*) of *Symbiodinium* within each coral core was measured prior to heat stress and at the end of heat stress using an Imaging PAM (Walz, Effeltrich, Germany). At the same time points, small tissue biopsies were taken from each core for genomic DNA extraction and quantification of the relative abundance of *Symbiodinium* clades using qPCR. Tissue samples were collected again after 3 months of recovery from bleaching to analyze changes in *Symbiodinium* community structure (i.e., symbiont shuffling). For additional details on sampling and data collection, see @Silverstein:2015er and @Cunning:2015ja.

@@

Measured values of *F~v~/F~m~* were significantly affected by the proportion of D1a in symbiont assemblages (Fig. 1), but this effect varied among coral species and between ambient and heat-stress conditions (three-way interaction, *F*=`r round(lmerTest::anova(bleachmod)$F[7], 3)`, *p*=`r round(lmerTest::anova(bleachmod)$P[7], 3)`). In all three coral species, *F~v~/F~m~* was negatively related to the proportion of D1a under ambient conditions, but positively related to the proportion of D1a under heat stress. The slope of each line in Fig. 1 (i.e., the difference in *F~v~/F~m~* when proportion D=1 and when proportion D=0) can be interpreted as the photochemical advantage of D1a compared to the other symbiont (Fig. 2). The advantage of D under ambient conditions (*A^D^~Amb~*) was significantly more negative (i.e., a greater disadvantage) in *O. faveolata* (`r round(AD[4,3], 3)`) compared to *M. cavernosa* (`r round(AD[6,3], 3)`, *p*=`r round(lst.pw$p.value[5], 3)`), and intermediate (`r round(AD[5,3], 3)`, but not significantly different, *p*>`r round(min(lst.pw$p.value[c(4,6)]), digits=1)`) in *S. siderea*. Under heat stress, the advantage of D1a (*A^D^~Heat~*) was greatest in *M. cavernosa* (`r round(AD[3,3], 3)`), intermediate in *S. siderea* (`r round(AD[2,3], 3)`), and lowest in *O. faveolata* (`r round(AD[1,3], 3)`, all pairwise comparisons, *p*<`r plyr::round_any(max(lst.pw$p.value[c(1,2,3)]), 0.001, ceiling)`).

@@

These experiments revealed significant differences in the photochemical efficiencies of co-occurring symbiont clades in mixed symbiont communities. At ambient temperatures, D1a symbionts had significantly lower *F~v~/F~m~* than the other symbiont clade within each coral host. Since ambient temperatures are not expected to cause photodamage, the reduced photochemical efficiency of D1a may be due to reduced nutrient sufficiency [@Parkhill:2001eu], greater pigment antenna size or association with PSI [@Suggett:2009p4201], greater excitation 'spillover' from PSII to PSI [@Slavov:2016kh], or other mechanisms of non-photochemical quenching [@Osmond:1994ua]. Indeed, inherent differences in photobiological traits produce taxonomic variation in *F~v~/F~m~* among broad microalgal classes [@Suggett:2009p4201] as well as different types of *Symbiodinium* [@Suggett:2008p960, @Suggett:2015dv]. Regardless of the underlying mechanisms, the reduced photochemical efficiency of D1a relative to other *Symbiodinium* *in hospite* suggests D1a has a reduced potential for PSII photochemistry. This disadvantage may explain why D-dominated corals tend to grow more slowly [@Little:2004p140], and why D1a may be displaced by other symbionts in the absence of stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364].

@@

This statistical result reveals predictors of symbiont shuffling but not necessarily the mechanisms underlying them. However, we speculate that if D1a performs better under at 32°C, it may be expelled at lower rates than co-occurring symbionts [@Silverstein:2017ep], or may even continue to proliferate. Furthermore, after heat stress is alleviated (but corals have bleached), amplified internal light environments characteristic of bleached corals [@Wangpraseurt:2012cu] may sustain the photochemical advantage of D1a, allowing it to proliferate at higher rates as the symbiont population regrows. Only once symbionts reach high densities may the internal microenvironment (e.g., darker, more nutrient-scarce conditions) again put D1a at a disadvantage, explaining why D may be gradually replaced over time in the absence of stress [@Thornhill:2006p3638]. Importantly, the rate of compositional change in the scenario outlined here would be proportional to the relative difference in the advantage of one symbiont over another, consistent with our observations across coral species.

>>>>

---
title: Symbiont shuffling linked to differential photochemical dynamics of *Symbiodinium*
  in three Caribbean reef corals
author:
- affiliation: MBE
  email: ross.cunning@gmail.com
  footnote: Corresponding Author
  name: Ross Cunning
- affiliation: MBE,MWK
  name: Rachel N. Silverstein
- affiliation: MBE
  name: Andrew C. Baker
address:
- address: Department of Marine Biology and Ecology, Rosenstiel School of Marine and
    Atmospheric Science, University of Miami, 4600 Rickenbacker Causeway, Miami, FL
    33149, USA
  code: MBE
- address: Miami Waterkeeper, 12568 N Kendall Drive, Miami, FL 33185, USA
  code: MWK
output:
  word_document: null
  rticles::elsevier_article:
    fig_caption: yes
csl: coral-reefs.csl
bibliography: library.bib
abstract: |
  Dynamic symbioses with functionally diverse dinoflagellate algae in the genus *Symbiodinium* may allow some reef corals to alter their phenotypes through 'symbiont shuffling', or changes in symbiont community composition. In particular, corals have been shown to become more bleaching-resistant by increasing the relative abundance of thermally tolerant *Symbiodinium* in clade D after bleaching. Despite the immediate relevance of this phenomenon to corals living in warming oceans -- and to interventions aimed at boosting coral resilience -- the mechanisms governing how, why, and when symbiont shuffling occurs are still poorly understood. Here, we performed controlled thermal bleaching and recovery experiments on three species of Caribbean reef-building corals, and show that the degree of symbiont shuffling is determined by (1) the difference in photochemical efficiency (*F~v~/F~m~*) of co-occurring symbionts under stress, and (2) the duration of stress exposure. The photochemical advantage of D1a (*S. trenchii*) under stress was greatest in *Montastraea cavernosa*, intermediate in *Siderastrea siderea*, and lowest in *Orbicella faveolata*, and correlated positively with the magnitude of shuffling toward D1a. In holobionts where D1a had less of an advantage under stress (i.e., only slightly higher *F~v~/F~m~*), longer periods of stress were required to elicit commensurate increases in D1a. These findings help elucidate general rules governing the process of symbiont shuffling in reef corals, which will help refine predictions of coral responses to environmental change and inform interventions to manipulate symbiont communities to enhance coral resilience.
---

```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_knit$set(root.dir = "..")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(lmerTest)
library(MASS)
```

@@

@@

Coral reef ecosystems are declining globally at an alarming rate, primarily due to mass coral bleaching caused by anthropogenic increases in sea surface temperature [@Hughes:2017bs]. Heat-induced bleaching, or the loss of corals' nutritional symbionts (*Symbiodinium* spp.; @Jokiel:1977p7353), often leads to coral mortality, which precipitates reef ecosystem degradation [@Baker:2008p3421].

@@

In both experiments, maximum photochemical efficiency (*F~v~/F~m~*) of *Symbiodinium* within each coral core was measured prior to heat stress and at the end of heat stress using an Imaging PAM fluorometer (Walz, Effeltrich, Germany). At the same time points, small tissue biopsies were taken from each core for genomic DNA extraction and quantification of the relative abundance of *Symbiodinium* clades using qPCR. Tissue samples were collected again after 3 months of recovery from bleaching to analyze changes in *Symbiodinium* community structure (i.e., symbiont shuffling). For additional details on sampling and data collection, see @Silverstein:2015er and @Cunning:2015ja.

@@

Measured values of *F~v~/F~m~* were significantly affected by the proportion of D1a in symbiont assemblages (Fig. 1), but this effect varied among coral species and between ambient and heat-stress conditions (three-way interaction, *F*=`r round(lmerTest::anova(bleachmod)$F[7], 3)`, *p*=`r round(lmerTest::anova(bleachmod)$P[7], 3)`). In all three coral species, a higher relative abundance of D1a led to lower *F~v~/F~m~* under ambient conditions, but higher *F~v~/F~m~* under heat stress. The slope of each line in Fig. 1 (i.e., the difference in *F~v~/F~m~* when proportion D=1 and when proportion D=0) can be interpreted as the photochemical advantage of D1a compared to the other symbiont (Fig. 2). The advantage of D under ambient conditions (*A^D^~Amb~*) was significantly more negative (i.e., a greater disadvantage) in *O. faveolata* (`r round(AD[4,3], 3)`) compared to *M. cavernosa* (`r round(AD[6,3], 3)`, *p*=`r round(lst.pw$p.value[5], 3)`), and intermediate (`r round(AD[5,3], 3)`, but not significantly different, *p*>`r round(min(lst.pw$p.value[c(4,6)]), digits=1)`) in *S. siderea*. Under heat stress, the advantage of D1a (*A^D^~Heat~*) was greatest in *M. cavernosa* (`r round(AD[3,3], 3)`), intermediate in *S. siderea* (`r round(AD[2,3], 3)`), and lowest in *O. faveolata* (`r round(AD[1,3], 3)`, all pairwise comparisons, *p*<`r plyr::round_any(max(lst.pw$p.value[c(1,2,3)]), 0.001, ceiling)`).

@@

These experiments revealed differences in the degree of symbiont shuffling in three coral species, which were strongly related to differences in the photochemical efficiencies of co-occurring symbionts. At ambient temperatures, D1a symbionts had significantly lower *F~v~/F~m~* than the other symbiont clade within each coral host. Since ambient temperatures are not expected to cause photodamage, the reduced photochemical efficiency of D1a may be due to reduced nutrient sufficiency [@Parkhill:2001eu], greater pigment antenna size or association with PSI [@Suggett:2009p4201], greater excitation 'spillover' from PSII to PSI [@Slavov:2016kh], or other mechanisms of non-photochemical quenching [@Osmond:1994ua]. Indeed, inherent differences in photobiological traits produce taxonomic variation in *F~v~/F~m~* among broad microalgal classes [@Suggett:2009p4201] as well as different types of *Symbiodinium* [@Suggett:2008p960, @Suggett:2015dv]. Regardless of the underlying mechanisms, the reduced photochemical efficiency of D1a relative to other *Symbiodinium* *in hospite* suggests D1a has a reduced potential for PSII photochemistry under ambient conditions. This disadvantage may explain why D-dominated corals tend to grow more slowly [@Little:2004p140], and why D1a may be displaced by other symbionts in the absence of stress [@Thornhill:2006p3638; @LaJeunesse:2009p7364].

@@

This statistical result reveals predictors of symbiont shuffling but not necessarily the mechanisms underlying them. However, we speculate that if D1a performs better under at 32°C, it may be expelled at lower rates than co-occurring symbionts [@Silverstein:2017ep], or may even continue to proliferate. Furthermore, after heat stress is alleviated (but corals have bleached), amplified internal light environments characteristic of bleached corals [@Wangpraseurt:2012cu] may sustain the photochemical advantage of D1a, allowing it to proliferate at higher rates as the symbiont population regrows. Only once symbionts reach high densities may the internal microenvironment (e.g., darker, more nutrient-scarce conditions) again put D1a at a disadvantage, explaining why clade D may be gradually replaced over time in the absence of stress [@Thornhill:2006p3638]. Importantly, the rate of compositional change in the scenario outlined here would be proportional to the relative difference in the advantage of one symbiont over another, consistent with our observations across coral species.
